---
title: "Dipping toes in Mixed Effects Modelling in R"
output:
  html_document:
    df_print: paged
---

R markdown tutorial - among thousands of them out there I (almost randomly) picked this one : https://r4ds.had.co.nz/r-markdown.html
```{r setup}
library(tidyverse)
library(viridis)
library(glue)
```

press Crtl+Alt+i and see what happens 

I am pretty much repeating what these web resources have to say about Mixed Effects Modelling:
https://ourcodingclub.github.io/2017/03/15/mixed-models.html
https://web.stanford.edu/class/psych252/section/Mixed_models_tutorial.html 
https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf


### grab the data and take a look
Crtl+Shift+Enter and see what happens 
```{r}
load("dragons.Rdata") 
head(dragons)
```
```{r}
str(dragons)
```


# First thing first, visualize your data! 

How's your outcome variable look like? 
```{r}
dragons %>% 
  ggplot(.,aes(testScore)) +
  geom_histogram(aes(y =..density..),alpha = .4)+
  geom_density(col="firebrick",size=1)+
  theme_bw()
```


```{r}
dragons %>% 
  ggplot(.,aes(x = bodyLength, y = testScore)) +
  geom_point() +
  geom_smooth(method = "lm")+
  theme_bw()
```

### Linear regression 
```{r}
basic.lm <- lm(testScore ~ bodyLength, data = dragons)
summary(basic.lm)
```

Is your model good looking?
```{r}
qqnorm(resid(basic.lm))
qqline(resid(basic.lm)) 
```


## Mixed effects modeling 

#### Use facet_wrap(~variable) or facet_grid(variable1~variable2)
Information we haven't utilize: Mountain range, site 

Again, know your data! 
```{r}
table(dragons$site,dragons$mountainRange)
```

```{r}
dragons <- dragons %>% 
  mutate(sampling_site = glue('{mountainRange}_{site}'))
head(dragons)
```


```{r, fig.width=10}
dragons %>% 
  ggplot(.,aes(bodyLength,testScore,col=site))+
  geom_point()+
  facet_wrap(~mountainRange,ncol=4)+
  theme_bw()+
  scale_color_viridis(discrete = T, option = "C")
```

Alternatively.
But I do prefer using facet_wrap()
```{r}
dragons %>% 
  ggplot(.,aes(bodyLength,testScore,col=mountainRange))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  theme_bw()+
  scale_color_viridis(discrete = T, option = "C")
```

### Linear mixed effect regression 
We want to tell the model how the data was generated/collected 
```{r}
library(lme4)
```

```{r}
mixed.intercept <- lmer(testScore ~ bodyLength + (1|mountainRange), data = dragons)
summary(mixed.intercept) 
```

A rough rule-of-thumb that the t is probably significant if it’s greater than 2.
Missing something? Hold one a second 

```{r}
library(lmerTest)
summary(mixed.intercept)
```

```{r}
qqnorm(resid(mixed.intercept))
qqline(resid(mixed.intercept)) 
```


```{r}
plot(mixed.intercept)
```

## Nested model
```{r}
mixed.nested<- lmer(testScore ~ bodyLength + (1|mountainRange)+(1|sampling_site), data = dragons)
summary(mixed.nested)
```

lmer function is smart enough to notice that sampling_site is nested under mountainRange,
but it's always a good practice to write your model like this 
```{r}
mixed.nested<- lmer(testScore ~ bodyLength + (1|mountainRange/sampling_site), data = dragons)
summary(mixed.nested)
```
```{r}
qqnorm(resid(mixed.nested))
qqline(resid(mixed.nested)) 
```
```{r}
plot(mixed.nested)
```

AIC=deviance+2∗(p+1) where p is # of parameters 
```{r}
anova(mixed.intercept,mixed.nested,refit=FALSE)
```
When comparing different random effects: 
`In lmer(), REML = TRUE (default), 
`In anova(), refit = FALSE

When comparing different random effects: 
`In lmer(), REML = FALSE , 
`In anova(), refit = TRUE (default)

## Plot the model fit
```{r}
dragons$fit <-predict(mixed.intercept)

dragons %>% 
  mutate(fit=predict(mixed.intercept)) %>% 
  ggplot(.,aes(bodyLength,testScore,group=mountainRange,col=mountainRange))+
  geom_point(alpha = .3,size=1.5)+
  theme_bw()+
  scale_color_viridis(discrete = T,)+
  geom_line(aes(y=fit),size = 1.2)+
  labs(x="How long is that bird",
      y="How smart is that bird",
      color = "Where is that bird from")+
  ggtitle("This is why we need to use mixed effects models")
```

## what do we make out of these results? 
So, the conclusion is: 
- you'll probably want to catch an ancient bird from mountain Julian, definitely not from mountain Bavarian or Southern. 

- Body length does NOT have an significant effect on test score. 

- there's something in that mountain varying both body length and test score...! 

# Sleep study 
The data is included in lme4 package 

```{r}
head(sleepstudy)
```

```{r}
str(sleepstudy) 
```


VISUALIZE
```{r, fig.width=10,fig.height=3.5}
sleepstudy %>% 
  ggplot(.,aes(Days,Reaction))+
  facet_wrap(~Subject,nrow = 2)+
  geom_point(shape = 21)+
  geom_smooth(method = "lm",se = F)+
  theme_bw()
```
```{r}
sleep.lm <- lm(Reaction ~ Days, sleepstudy)
summary(sleep.lm)
```

```{r}
sleep.intercept <- lmer(Reaction ~ Days + (1| Subject), sleepstudy)
summary(sleep.intercept)
```

NOTE: in many cases, models may fail to converge (meaning the model is sh**) when you are trying to fit random slope, because fitting random slope requires a large sample size. 

Don't worry... RED warning message will tell you when it fails to converge. 

Harrison et al., (2018): "fitting the most complex mixed effects structrue allowed by your data by" 

(i) fitting random slopes but removing the correlation between interecepts and slopes

(ii) fitting no random slopes at all but accepting that this inflates the Type I error rate
```{r}
sleep.slope <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
summary(sleep.intercept)
```


```{r}
sleep.slope.no.correlation <- lmer(Reaction ~ Days + ( Days || Subject), sleepstudy)
summary(sleep.slope.no.correlation)
```


```{r, fig.width=10}
p1 <- sleepstudy %>% 
  mutate(pred_dist = fitted(sleep.intercept)) %>% 
  ggplot(.,aes(Days, pred_dist, col = Subject))+
  theme_classic()+
  geom_line(size = 1)+
  scale_color_viridis(discrete = T,option = "C")

p2 <- sleepstudy %>% 
  mutate(pred_dist = fitted(sleep.slope)) %>% 
  ggplot(.,aes(Days, pred_dist, col = Subject))+
  theme_classic()+
  geom_line(size = 1)+
  scale_color_viridis(discrete = T,option = "C")

p3 <- sleepstudy %>% 
  mutate(pred_dist = fitted(sleep.slope.no.correlation)) %>% 
  ggplot(.,aes(Days, pred_dist, col = Subject))+
  theme_classic()+
  geom_line(size = 1)+
  scale_color_viridis(discrete = T,option = "C")

gridExtra::grid.arrange(p1,p2,p3,nrow=1)

```

```{r}
anova(sleep.intercept,sleep.slope, sleep.slope.no.correlation, refit = FALSE)
```

# Wheels lol
```{r}
df <- read.csv("wheels.csv")
head(df)
```

```{r}
df1 <- df %>% 
  gather(day,rev,2:5) %>% 
  mutate(day = as.numeric(str_sub(day,4)))
head(df1)
```

```{r}
df1 %>% 
  ggplot(.,aes(rev))+
  geom_histogram()
```


```{r}
df1 %>% 
  ggplot(.,aes(day,rev, group = id))+
  geom_line(color = "grey")+
  geom_line(data = df1 %>% 
              group_by(strain,day) %>% 
              summarize(avg = mean(rev,na.rm = T)),
            aes(day,avg,color=strain,group = strain))+
  facet_wrap(~strain)+
  theme_classic()
```

```{r}
mod0 <- lmer(rev ~ 1 + (day||id), data = df1, REML = F)

mod1 <- lmer(rev ~ strain*day + (day||id), data = df1, REML = F)

summary(mod1)
```

```{r}
plot(mod1)
```


```{r}
anova(mod0,mod1)
```

```{r}
library(MCMCglmm)
```

```{r}
df2 <- df1 %>% 
  mutate(id = as.character(id))


mcmc.mod <- MCMCglmm(rev ~ day*strain, 
                      random = ~idh(day):id,
                      family = "gaussian",
                      data = df2,
                      verbose = F)

summary(mcmc.mod)
```

```{r}
plot(mcmc.mod$Sol)
```

